Supplementary figures\nCapturing regeneration in models of forest dynamics

### 2023-04-28 manuscript/supplementaryRW.Rmd
xComp report updates from Tobi

Tobias Mette, LWF, 24.05.2022

@@

The forest growth model xcomp was developed to simulate gap dynamics of
monospecific stands with a strong emphasis on position-dependent
individual tree competition. The model consists of three units: (1) the
main growth phase which determines the growth and mortality of each
individual tree above 5 cm dbh, (2) the old-growth phase where
individual trees exceeding a certain dbh-limit decline in vitality and
eventually die, and (3) the regeneration phase where seedling growth and
mortality is modelled depending on light availability. The primary focus
of the model was to establish a set of robust algorithms that can be
flexibly calibrated for different species attributes. The model was used
to simulate single-species stand development cycles for Nothofagus
pumilio in Patagonia (Mette 2014), for European beech in NE-Bavaria
(Mette et al. 2016) and a Norway spruce stand with different mortality
algorithms (Bugmann et al. 2019). For the current article the model was
(4) connected to a climate sensitive species distribution model to
modify growth and mortality in dependence of the site conditions, and
(5) extended from single to multi-species simulations.

The most complete reference resource for the xComp model and
regeneration module is:

-

@@

@@

@@

@@

@@

The simulation was carried out according to the protocol by Bugmann et
al. (2022). I.e. the simulation is initiated for mixed species stands of
1 ha size at 200 sites of with given climate and soil conditions. The
initial species mix consists of up to 10 species of equal shares with an
initial root mean square diameter dq of 8 cm. From the suggested species
spectrum Alnus glutinosa and Fraxinus excelsior were not included in the
species mix. Quercus spec was split into Q. petraea and Q. pubescens,
but united in the validation to meet the protocol. Xcomp simulates
forest growth in 5 year intervals. Data are validated after having
reached a climate dependent equilibrium -- which is typically the case
after 750 years (= 150 simulation periods). Starting with year 760, data
are read out for 200 10 year intervals (until year 2750, simulation
period 550).

#### Model details

(1) The main growth phase is determined by the processes growth and
    mortality. Mortality is determined by the Reineke exponent rke and
    the stand density sdi and calibrated over a range of 8 cm to 50 cm
    dq (= root mean square diameter) so that stem number N equals: N =
    sdi*(dq/25)\^rke. A tree's resource acquisition area res (m2) is
    approximated as res = 1e4/ (sdi*(dbh/25)\^rke). As the Reineke
    exponent is fixed to 1.6, species with higher/ lower sdi have a
    lower/ higher resource acquisition area for a given dbh. Potential
    dbh growth dd is calculated by multiplying a reference dbh growth
    with a site specific temperature dependent modifier. The individual
    dbh growth in the stand is further modified by the neighbourhood
    competition.

@@

(3) The regeneration phase allows seedling development on a 1x1 m grid
    over the entire stand. First, a light availability matrix (1x1 m) is
    calculated from a cumulative crown shadow which decreases with
    increasing distance to the tree center. Next, regeneration matrices
    (1x1 m) are created for each of the stand's species. In the initial
    period, the regeneration matrix of each species is empty, but at the
    end of each period, a light dependent dbh growth and mortality is
    calculated for each 1m raster cell. The dbh growth is scaled so that
    under full light 5 cm dbh is reached in a defined regeneration
    period (regnPer) that depends on the site conditions. The default
    value is set to 15 years and equal for all species. Dbh growth is
    modified by the light availability and the species' regeneration
    light demand. This is a decisive difference between the species and
    favours shadow-tolerant trees like fir and beech in the absence of
    large gaps. However, each raster cell is afflicted with a species
    bonus or malus that provides local variation to the global
    regeneration parameters. Mortality in the regeneration raster occurs
    as a random process, the likelyhood of which depends on the climate.

@@

The processes that determine the species composition and turnover in
xcomp can be explained best by an example. In Fig. 1, I displayed the
basal area shares resulting from the simulation of site 434 over a time
span of 2750 years. Tab. 1 shows the species calibration parameters for
the site. It is a very favourable site with modelled species prevalences
of 0.93 (beech), 0.86/7 (spruce, maple), 0.81/2 (fir, ped. oak), 0.76/4
(hornbeam, lim, birch), and 0.70 (pine). Starting with more or less
equal basal area shares and 8 cm dq (root mean square dbh), all species
increase their basal area in the first 50 years. Thereafter increased
competition and starting old growth leads to a decline -- first, in all
of the species but spruce, fir and beech, but after 100 years also in
spruce. Only in fir and beech regeneration balances the loss of older
trees. The result is an equilibrium of a dense fir and beech forest with
high basal area where more light demanding species regenerate too little
to contribute significantly.

#### Discussion

xcomp is a very deterministic model where except for mortality and a
random species regeneration malus/ bonus on the regeneration raster.
This leads to equilibria of one or two species. What species depends
strongly on the light demand of the regeneration which favours fir and
beech in the absence of large gaps creating disturbances. The second
most important parameter is the site suitability which is estimated by
the modelled species prevalence and affects growth and thereby
competitivity. As it is the first time that xcomp is used in a multiple
species application there is room for improvements. The automation of a
robust calibration of the main growth phase is a precondition to
investigate how for instance larger disturbances, a weighted
regeneration potential oriented at the main stand, higher or selective
mortality of the regeneration affect the species composition and
turnover.

### references

1.  Bugmann, H., Seidl, R., Hartig, F., Bohn, F., Brůna, J., Cailleret,
    M., ... & Reyer, C. P. (2019). Tree mortality submodels drive
    simulated long‐term forest dynamics: Assessing 15 models from the
    stand to global scale. Ecosphere, 10(2), e02616.
2.  Karger, D. N., Conrad, O., Böhner, J., Kawohl, T., Kreft, H.,
    Soria-Auza, R. W., ... & Kessler, M. (2017). Climatologies at high
    resolution for the earth's land surface areas. Scientific data,
    4(1), 1-20.
3.  Mauri, A., Strona, G., & San-Miguel-Ayanz, J. (2017). EU-Forest, a
    high-resolution tree occurrence dataset for Europe. Scientific data,
    4(1), 1-8.
4.  Mette, T. (2014) Using Reineke to self-calibrate mortality in
    individual tree based forest growth modelling. Beiträge zur
    Forstwissenschaftlichen Tagung, Dresden, 17.-20.09.2014
5.  Mette, T., Falk, W., Blaschke, M., Förster, B., Walentowski, H.
    (2016). Modelling natural forest dynamics in individual tree based
    forest growth simulators. Beiträge zur Forstwissenschaftlichen
    Tagung, Freiburg, 26.-29.09.2016
6.  Thurm, E. A., Hernandez, L., Baltensweiler, A., Ayan, S.,
    Rasztovits, E., Bielak, K., ... & Falk, W. (2018). Alternative tree
    species under climate warming in managed European forests. Forest
    Ecology and Management, 430, 485-497.

>>>>

Tobias Mette, LWF.

@@

The forest growth model xcomp was developed to simulate gap dynamics of
monospecific stands with a strong emphasis on position-dependent
individual tree competition. The model consists of three units: (1) the
main growth phase which determines the growth and mortality of each
individual tree above 5 cm dbh, (2) the old-growth phase where
individual trees exceeding a certain dbh-limit decline in vitality and
eventually die, and (3) the regeneration phase where seedling growth and
mortality is modelled depending on light availability. The primary focus
of the model was to establish a set of robust algorithms that can be
flexibly calibrated for different species attributes. The model was used
to simulate single-species stand development cycles for Nothofagus
pumilio in Patagonia (Mette 2014), for European beech in NE-Bavaria
(Mette et al. 2016) and a Norway spruce stand with different mortality
algorithms (Bugmann et al. 2019). For the current article the model was
(4) connected to a climate sensitive species distribution model to
modify growth and mortality in dependence of the site conditions, and
(5) extended from single to multi-species simulations (c.f. Section
1.6.7).

The most complete reference resource for the xComp model and
regeneration module are:

-   Mette, T. (2014). xcomp Tutorial (v2014_08). Technical Report, TUM.
    DOI: 10.13140/RG.2.2.32524.59529

-   Mette, T. (2014) Using Reineke to self-calibrate mortality in
    individual tree based forest growth modelling. Beiträge zur
    Forstwissenschaftlichen Tagung, Dresden,
    17.-20.09.2014. 10.13140/RG.2.2.11553.07527

@@

The species parameterisation in xcomp is sensitive to long-term climate,
i.e. 30-year averages of summer temperature, summer precipitation and
winter temperature. Reactions to daily, monthly or periodic extremes are
not implemented as deterministic but as stochastically processes
functionally dependent on the long-term climate. The climate sensitivity
of growth, mortality and regeneration is described in Section 1.6.7.

@@

Soil moisture class MOIST was used to modify the climate sensitivity via
summer precipitation PJJA by PJJA(mod) = PJJA+15\*MOIST-45 (in mm).

@@

Topography is no input parameter in xcomp. It may only effect the
simulation results indirectly through climate or soil parameters.

@@

From the suggested species spectrum *Alnus glutinosa* and *Fraxinus
excelsior* were not included in the species mix. *Quercus spec* was
split into *Q. petraea* and *Q. pubescens*, but united in the validation
to meet the protocol.

@@

The simulation was carried out according to the protocol by Bugmann et
al. (2022). I.e. the simulation is initiated for mixed species stands of
1 ha size at 200 sites of with given climate and soil conditions. The
initial species mix consists of up to 10 species of equal shares with an
initial root mean square diameter dq of 8 cm. Xcomp simulates forest
growth in 5 year intervals. Data are validated after having reached a
climate dependent equilibrium -- which is typically the case after 750
years (= 150 simulation periods). Starting with year 760, data are read
out for 200 10 year intervals (until year 2750, simulation period 550).
Further simulation relevant model details:

(1)   The main growth phase is determined by the processes growth and
mortality. Mortality is determined by the Reineke exponent rke and the
stand density sdi and calibrated over a range of 8 cm to 50 cm dq (=
root mean square diameter) so that stem number N equals: N =
sdi\*(dq/25)\^rke. A tree\'s resource acquisition area res (m2) is
approximated as res = 1e4/ (sdi\*(dbh/25)\^rke). As the Reineke exponent
is fixed to ‑1.6, species with higher/ lower sdi have a lower/ higher
resource acquisition area for a given dbh. Potential dbh growth dd is
calculated by multiplying a reference dbh growth with a site specific
temperature dependent modifier. The individual dbh growth in the stand
is further modified by the neighbourhood competition.

@@

(3)   The regeneration phase allows seedling development on a 1x1 m grid
over the entire stand. First, a light availability matrix (1x1 m) is
calculated from a cumulative crown shadow which decreases with
increasing distance to the tree center. Next, regeneration matrices
(1x1 m) are created for each of the stand\'s species. In the initial
period, the regeneration matrix of each species is empty, but at the end
of each period, a light dependent dbh growth and mortality is calculated
for each 1m raster cell. The dbh growth is scaled so that under full
light 5 cm dbh is reached in a defined regeneration period (regnPer)
that depends on the site conditions. The default value is set to 15
years and equal for all species. Dbh growth is modified by the light
availability and the species\' regeneration light demand. This is a
decisive difference between the species and favours shadow-tolerant
trees like fir and beech in the absence of large gaps. However, each
raster cell is afflicted with a species bonus or malus that provides
local variation to the global regeneration parameters. Mortality in the
regeneration raster occurs as a random process, the likelyhood of which
depends on the climate.

@@

Since the simulation output tables contain all biometric variables on an
individual tree basis, all required outputs can be derived directly from
the output tables. Being a very deterministic model what typically
happens in xcomp is that the equilibria consist of one or two species.
Shade tolerance favours fir and beech in the absence of large gaps
creating disturbances. The second most important parameter is the site
suitability which is estimated by the modelled species prevalence and
affects growth and thereby competitivity. As it is the first time that
xcomp is used in a multiple species application there is room for
improvements. The automation of a robust calibration of the main growth
phase is a precondition to investigate how for instance larger
disturbances, a weighted regeneration potential oriented at the main
stand, higher or selective mortality of the regeneration affect the
species composition and turnover.

1.  Bugmann, H., Seidl, R., Hartig, F., Bohn, F., Brůna, J., Cailleret,
    M., \... & Reyer, C. P. (2019). Tree mortality submodels drive
    simulated long‐term forest dynamics: Assessing 15 models from the
    stand to global scale. Ecosphere, 10(2), e02616.
2.  Karger, D. N., Conrad, O., Böhner, J., Kawohl, T., Kreft, H.,
    Soria-Auza, R. W., \... & Kessler, M. (2017). Climatologies at high
    resolution for the earth\'s land surface areas. Scientific data,
    4(1), 1-20.
3.  Mauri, A., Strona, G., & San-Miguel-Ayanz, J. (2017). EU-Forest, a
    high-resolution tree occurrence dataset for Europe. Scientific data,
    4(1), 1-8.
4.  Mette, T., Falk, W., Blaschke, M., Förster, B., Walentowski, H.
    (2016). Modelling natural forest dynamics in individual tree based
    forest growth simulators. Beiträge zur Forstwissenschaftlichen
    Tagung, Freiburg, 26.-29.09.2016. 10.13140/RG.2.2.29169.15201
5.  Thurm, E. A., Hernandez, L., Baltensweiler, A., Ayan, S.,
    Rasztovits, E., Bielak, K., \... & Falk, W. (2018). Alternative tree
    species under climate warming in managed European forests. Forest
    Ecology and Management, 430, 485-497.
